Formalizing Theoretical Commutative Algebra in Coq: \\ Nakayama's Lemma

### 2022-10-19 nakayama/paper.tex
Started introduction and added minor updates to preamble and references

\newtheorem*{theorem}{Theorem}

@@

@@

For the proof of Nakayama's lemma, it too follows the standard, informal
mathematical argument. Do induction on the number of vectors needed to generate
the module \(M\). The base case where \(M\) is generated by 0 vectors is by
definition true since an empty linear combination is conventionally taken to be
the zero vector. The inductive case where \(M\) is generated by the vectors
\(b_{1}, \dots, b_{m}, b_{m+1}\). By assumption \(M = \mathfrak{m} M\) with
\(b_{1}\in M\). Then \(b_{1}\in\mathfrak{m} M\) meaning that by our lemma,
there exists a linear combination for \(b_{1}\), say
\[b_{1} = u_{1} b_{1} + \cdots + u_{m+1} b_{m+1}\]
for some \(u_{1}, \dots, u_{m+1}\in\mathfrak{m}\). Collect the \(b_{1}\) terms
on the left-hand side of the equation to get that
\[(1 - u_{1}) b_{1} = u_{2} b_{2} + \cdots + u_{m+1} b_{m+1}\text{.}\]
As mentioned when defining maximal ideals, \(\mathfrak{m}\) can only contain
non-units. Namely, \(u_{1}\) must be a non-unit. Then we had already proven
that \(1 - u_{1}\) is a unit as we are in a local ring. Let \(v_{1}\) be the
multiplicative inverse of \(1 - u_{1}\), and multiply it on both sides of the
equation.
\begin{align*}
  v_{1} (1 - u_{1}) b_{1}
    & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
  1 b_{1} & = \\
  b_{1} & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
\end{align*}
We have thus found that one of the basis vectors is not needed to generate this
module. Using the induction hypothesis, we have that \(M = 0\). Showing that
the induction hypothesis holds in Coq takes more work than in an informal
proof, but this extra work is just a lot of bookkeeping.
\end{document}

>>>>

\input{preamble.tex}

@@

\title{Formalizing Theoretical Commutative Algebra in Coq: \\ Nakayama's Lemma}

%\thanks{}
\date{}

%\author{}
%\address{}
%\email{}

\maketitle

%\begin{abstract}
%	Abstract here
%\end{abstract}

\section{Introduction}

A number of structures from abstract algebra have been formalized using the proof assistant Coq \cite{the_coq_development_team_2019_3476303}.  
That said, for higher-level algebra, there currently appears to be a computational emphasis. 
For instance, the commutative-algebraic notion of a Gr\"obner basis, a special type of generating set for an ideal in a ring of polynomials, and which has cryptographic applications, has gathered significant attention (e.g., see \cite{thery-buchberger}).

The mathematical field of commutative algebra not only has further applications to cryptography, but also in other areas of science and technology such as data science and the study of the human genome.
Commutative algebra stems from the study of polynomial equations, and now centers around commutative rings--rings whose multiplication satisfies the commutative law, ideals of these rings, and modules over them.  It has powerful connections with other areas of theoretical mathematics including number theory and algebraic geometry.

We create a formal proof of Nakayama's lemma, a classical theorem in commutative algebra, in Coq. 
%
% ``lemma'' appearing in its name 
%\footnote{},
%
In doing so, we formalize, from scratch, certain algebraic structures that are essential to higher-level algebra, including local rings and modules over commutative rings. 
%
The notion of a module over a ring is an extension of the linear-algebraic notion of a vector space over a field, ubiquitous in mathematics and its applications. 
Nakayama's lemma exhibits one way that a finitely generated module over a commutative ring behaves like a vector space over a field.

\begin{nak*}
Let $R$ be a commutative local ring with maximal ideal $\mathfrak{m}$, and suppose that $M$ is a finitely generated $R$-module.  If $M = \mathfrak{m} M$, then $M = 0$.
\end{nak*}

\section{Mathematical Basis and Motivation}

\section{Formalization}

@@

For the proof of Nakayama's lemma, it too follows the standard, informal
mathematical argument. Do induction on the number of vectors needed to generate
the module \(M\). The base case where \(M\) is generated by 0 vectors is by
definition true since an empty linear combination is conventionally taken to be
the zero vector. The inductive case where \(M\) is generated by the vectors
\(b_{1}, \dots, b_{m}, b_{m+1}\). By assumption \(M = \mathfrak{m} M\) with
\(b_{1}\in M\). Then \(b_{1}\in\mathfrak{m} M\) meaning that by our lemma,
there exists a linear combination for \(b_{1}\), say
\[b_{1} = u_{1} b_{1} + \cdots + u_{m+1} b_{m+1}\]
for some \(u_{1}, \dots, u_{m+1}\in\mathfrak{m}\). Collect the \(b_{1}\) terms
on the left-hand side of the equation to get that
\[(1 - u_{1}) b_{1} = u_{2} b_{2} + \cdots + u_{m+1} b_{m+1}\text{.}\]
As mentioned when defining maximal ideals, \(\mathfrak{m}\) can only contain
non-units. Namely, \(u_{1}\) must be a non-unit. Then we had already proven
that \(1 - u_{1}\) is a unit as we are in a local ring. Let \(v_{1}\) be the
multiplicative inverse of \(1 - u_{1}\), and multiply it on both sides of the
equation.
\begin{align*}
  v_{1} (1 - u_{1}) b_{1}
    & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
  1 b_{1} & = \\
  b_{1} & = v_{1} u_{2} b_{2} + \cdots + v_{1} u_{m+1} b_{m+1} \\
\end{align*}
We have thus found that one of the basis vectors is not needed to generate this
module. Using the induction hypothesis, we have that \(M = 0\). Showing that
the induction hypothesis holds in Coq takes more work than in an informal
proof, but this extra work is just a lot of bookkeeping.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
